---
title: "stepwise model selection"

---

# Model selection procedure

Variable descriptions:

- set_ID: unique set identifier                 
- trip_ID: unique trip identifier (not used - missing data & errors)                
- fisher_ID: unique identifier for the captain
- year: calendar year (2012-2016) 
- fishing_date_mod: day-month-year extracted from the dttm 
- month_num: numeric month extracted from fishing_date              
- set_time_mod_corr: time the gear was set (HH:MM:SS)
- latitude_set_dd_mod: latitude the gear was set in decimal degrees  
- longitude_set_dd_mod: longitude where gear was set
- all_tuna_likes_kg: total kg of all tuna and tuna like species         
- non_tuna_retained_kg: total kg of other retained species  
- turtles_cnt: sea turtles (all species combined, total count)                
- dolphins_cnt: (all species combined, total count)             
- fishing_method: Three categories in local names ("Hojo/oojo", "Rach Jal", "Jaal Bodo")         
- fishing_method_2: Simplified to two categories: Surface (Hojo/oojo + Rach Jal) and Bottom gillnets (Jaal Bodo)       
- no_crew: Number of crew onboard                
- no_panel: Number of gillnet panels              
- length_panel_m: length of each panel in meters         
- width_net_m: width of each panel in meters   
- net_area_m2: Effort offset (no_panel * length_panel_m * width_net_m)

```{r setup, include=FALSE}

library(tidyverse)
library(here)
library(pillar)
library(mgcv)
library(MuMIn)
library(broom)
library(glmmTMB)

knitr::opts_chunk$set(echo = TRUE)
```

# prep data
```{r}
glimpse(model.df.wide) # 5,599 sets

model.df.v2 <- model.df.wide %>% 
  filter(fisher_ID != 14) %>% # drop fisher 14 (they drop out anyways due to many NA variables)
  filter(!month_num %in% c(6,7)) %>% # remove those closed season sets from fisher 4 (experimental trials, not normal fishing operations)
  mutate(across(c(fisher_ID, fishing_method, fishing_method_2), as.factor)) 
# final model df has 5,502 sets 

# format time as hour.decimal
class(model.df.v2$set_time_mod_corr)
model.df.v2$set_time_mod_corr_num <- as.numeric(model.df.v2$set_time_mod_corr) / 3600
model.df.v2$set_time_mod_corr_num <- round(model.df.v2$set_time_mod_corr_num * 6) / 6
model.df.v2$set_time_mod_corr_num <- round(model.df.v2$set_time_mod_corr_num, 1)

```

# recode month as chronological month of the fishing season

```{r recode month}
model.df.v2 <- model.df.v2 %>%
  mutate(
    cal_month = month(fishing_date_mod),          # 1 = Jan, ..., 12 = Dec
    season_month = case_when(
      cal_month >= 8  ~ cal_month - 7,     # Aug–Dec 1–5
      cal_month <= 5  ~ cal_month + 5,     # Jan–May 6–10
      TRUE            ~ NA_real_           # Jun–Jul (no fishing)
    )
  )
```

# Dolphins

```{r Dolphins}

##_______________________ DOLPHINS
dols.global <- gam(dolphins_cnt ~ year + s(all_tuna_likes_kg) + s(non_tuna_retained_kg) + s(season_month, k=3) + s(latitude_set_dd_mod, longitude_set_dd_mod) + s(set_time_mod_corr_num) + s(no_crew) + fishing_method_2 + s(fisher_ID, bs = "re"), 
                   offset = log(net_area_m2), data = model.df.v2, family = tw(), na.action="na.omit", method="REML", select=TRUE)
AIC(dols.global) 
summary(dols.global) 

##------------- check negative binomial
dols.nb <- update(dols.global, family = nb())
AIC(dols.nb, dols.global) # tweedie is better

##-------------------- Compare to fixed effect
dols.global.fe <- update(dols.global, .~. -s(fisher_ID, bs = "re") + fisher_ID)
AIC(dols.global.fe) # only ~0.3 AIC points worse with fixed effect captain term compared to RE
summary(dols.global.fe) # SEs for captain coefficients are huge

##----------------- comapre to the local names fishign method
dols.fe.2 <- update(dols.global.fe, .~. - fishing_method_2 + fishing_method)
AIC(dols.fe.2) # worse and still not significant

##---------------- Simplify models and adjust parameterization
dols.fe.3 <- update(dols.global.fe, .~. - s(set_time_mod_corr_num) - fishing_method_2)
# continue simplifying and checking

##-----------------Compare to fixed effect with sum contrast coding & paraPenn on captian term
# set up paraPen
NoCoefs <- length(unique(model.df.v2$fisher_ID))-1 
pp <- list(fisher_ID = list(rank = NoCoefs, diag(NoCoefs)))
# set up contrast coding
contrasts(model.df.v2$fisher_ID) <- contr.sum

dols.fe.best <- gam(dolphins_cnt ~ year + s(all_tuna_likes_kg) + s(latitude_set_dd_mod, longitude_set_dd_mod, k=50) + fisher_ID,
    offset = log(net_area_m2), data = model.df.v2, family = tw(), na.action="na.omit", method="REML", paraPen=pp, select=TRUE)

##------------------- compare final model minus captain effect 
dols.test <- update(dols.fe.best, .~. -fisher_ID)

```

# Sea turtles
```{r sea turtles}

##________________________SEA TURTLES
turts.global <- gam(turtles_cnt  ~ year + s(all_tuna_likes_kg) + s(non_tuna_retained_kg) + s(season_month, k=3) + s(latitude_set_dd_mod, longitude_set_dd_mod, k=50) + s(set_time_mod_corr_num) + s(no_crew) + fishing_method_2 + s(fisher_ID, bs = "re"), 
  offset = log(net_area_m2), data = model.df.v2, family = tw(), na.action="na.omit", method="REML", select=TRUE)
AIC(turts.global) 
summary(turts.global) 

##---------------compare to negative binomial
turts.nb <- update(turts.global, family = nb())
AIC(turts.nb, turts.global) # tweedie much better

##-------------------- Compare to fixed effect

turts.global.fe <- update(turts.global, .~. - s(fisher_ID, bs = "re") + fisher_ID)
AIC(turts.global.fe) # beats random effect
summary(turts.global.fe)

# Basic GAM diagnostics
gam.check(mod)
# Concurvity
concurvity(mod, full = TRUE)
# Smooth term plots
plot(mod, pages = 1, residuals = TRUE, shade = TRUE)
# DHARMa residual diagnostics
library(DHARMa)
sim <- simulateResiduals(mod)
plot(sim)
testDispersion(sim)
testZeroInflation(sim)


##---------------- Simplify models and adjust parameterization

turts.fe.2 <- update(turts.global.fe, .~. -s(non_tuna_retained_kg))

turts.fe.3 <- udpate(turts.fe.2, .~. -fishing_method_2 + fishing_method) # compare the 3 category local name factor to the simplified factor with 2 levels

# continue subtracting terms and checking AIC/summary/diagnostics
# adjust ks where necessary
# check smooth edfs to see if relationship looks linear
# etc


##-----------------Compare to fixed effect with sum contrast coding & paraPenn on captian term
# set up paraPen
NoCoefs <- length(unique(model.df.v2$fisher_ID))-1 
pp <- list(fisher_ID = list(rank = NoCoefs, diag(NoCoefs)))
# set up contrast coding
contrasts(model.df.v2$fisher_ID) <- contr.sum

turts.fe.best <- gam(turtles_cnt ~ year + s(all_tuna_likes_kg) + s(season_month, k = 3) + 
    s(latitude_set_dd_mod, longitude_set_dd_mod) + s(no_crew) + fisher_ID,
    offset = log(net_area_m2), data = model.df.v2, family = tw(), na.action="na.omit", method="REML", select=TRUE)
# no evidence the paraPen improved the sea turtles model

##------------------- compare final model minus captain effect 
turts.test <- update(turts.fe.best, .~. - fisher_ID)
AIC(turts.test) # worse wihtout captain

# check model diagnostics
```

# save best model outputs
model summaries for results table
GAM plots
```{r save outputs}

library(openxlsx) 
library(mgcViz)
library(pixiedust)
library(gratia) 

#mod <- dols.fe.best
mod <- turts.fe.best

summary.mod <- summary(mod)
param <- rownames(summary.mod$p.table)
coeffs.out <- cbind(param, as.data.frame(summary.mod$p.table, rownames=NULL))
mod.out <- openxlsx::createWorkbook()
mod.glance <-  broom::glance(mod)
mod.sum <- pixiedust::dust(mod, glance_foot=TRUE) %>%
  sprinkle(col=2:4, round=4) %>%
  sprinkle(cols=5, fn=quote(pvalString(value))) %>%
  sprinkle(rows = 1, border = "top") %>%
  sprinkle(cols = c(2, 5), round = 2, na_string = "",
           part = "foot") %>%
  sprinkle(rows = 1, border = "top", part = "foot") %>%
  sprinkle_print_method("console")
mod.out <- createWorkbook()
addWorksheet(mod.out, "gam_glance")
addWorksheet(mod.out, "gam_summ")
addWorksheet(mod.out, "coeffs_full")
writeData(mod.out, sheet = "gam_glance", x=mod.glance)
writeData(mod.out, sheet = "gam_summ", x=mod.sum)
writeData(mod.out, sheet = "coeffs_full", x=coeffs.out)


# Diagnostics plots
appraise(dols.fe.best)
appraise(turts.fe.best)

# smooth plots 
smooths.dols <- gratia::draw(dols.fe.best, residuals = TRUE, rug = FALSE) & theme_bw(base_size = 12) 
smooths.turts <- gratia::draw(turts.fe.best, residuals = TRUE, rug = FALSE) & theme_bw(base_size = 12) 

```
